Belowground sink limitation alters growth and carbon balance of *Eucalyptus* seedlings

### 2016-01-07 manuscript/manuscript_pve.Rmd
pve manuscript to tree physliogy update

\
RUNNING HEAD:

@@

\
In woody species, the coordination of *A*~n~ and growth has been studied with manipulations of *C* source activity. Examples included elevated CO~2~ experiments, for example FACE [reviewed in @ainsworth2005have], and partial defoliation experiments. Elevated CO~2~ has been shown to increase *A*~n~ [@drake1997more; @ainsworth2007response] and across four FACE experiments this resulted in a stimulation of 23 % in forest biomass production [@norby2005forest]. Evidence from a wide range of elevated CO~2~ experiments, however, also reveals that even with an average photosynthetic enhancement of over 30 %, the biomass growth rate only increases by around 10 % [@kirschbaum2011does]. In partial defoliation experiments, increases in *A*~n~ of the remaining foliage are commonly shown, yet are attributed to various mechanisms, including reduction in end product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end], increased stomatal conductance [@layne1995end], enhanced leaf nutrient status [@turnbull2007increased] and regulatory sugar signaling [@eyles2013whole]. However, increases in *A*~n~ in defoliation experiments did not always produce increased growth due to reductions in meristem sink strength [@palacio2012fast], *C* limitation to mycorrhizal colonization [@markkola2004defoliation], or an overall decrease in whole plant *C* gain [@ovaska1993a]. These manipulations of C source activity expose unresolved issues with how changes in *A*~n~ do not always infer proportional responses in growth.

\
Alternatively, manipulating plant tissue *C* sinks is often used to investigate the correlation between *A*~n~ and growth. This is because metabolic signaling networks, relaying information on *C* and *N* status of different tissues, can regulate photosynthetic activity [@paul2001sink]. If sink inhibition of *A*~n~ occurs, a close coordination between declines in *A*~n~ and growth should be expected. Whether photosynthetic down regulation is evident in woody species has been tested through fruit removal and phloem girdling. In these studies, down regulation of *A*~n~ was frequently correlated to carbohydrate accumulation resulting from reduced tissue sink strength  [@iglesias2002regulation;@urban2007girdling; @haouari2013fruit]. However, reductions in *A*~n~ were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic] and stomatal limitation [@li2005photosynthesis]. These mixed results are not surprising as we still know little about the balance between assimilation, storage and growth across temporal scales in plants [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to the occurrence of large fruiting sinks, they tell us little about source-sink coordination in typical growing conditions for woody species.

\
An alternative experimental approach is to reduce belowground *C* sink strength in tree seedlings by manipulating rooting volume, by varying the container size [@arp1991effects; @nesmith1998effect; @poorter2012pot]. Possible advantages of this approach are that it allows a large range of treatment levels, can be easily compared to naturally planted seedlings and may mimic natural conditions as seedlings compete for space or reach bedrock. Seedlings undergo many physiological and morphological changes in response to rooting volume, including biomass partitioning, *A*~n~, water relations, nutrient uptake and respiration [@nesmith1998effect;@poorter2012pot and references therein]. Inadequate rooting volume may decrease *C* sink strength by progressively restricting root growth in growing plants [@thomas1991root].  Container size studies frequently exhibit photosynthetic down-regulation, likely as a result of sink limitation [@arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. A meta-analysis by Poorter et al. [-@poorter2012pot] concluded that *A*~n~ is the process likely to be the strongest affected by pot size and may best explain the effects on biomass seen in the large number of studies where containers are used. This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower *C* exchange and assimilate export rates [@robbins1988effect]. However, evidence in support for a trade-off between *C* storage and growth in trees is, to date, inconclusive [@palacio2014does]. Based on these previous studies, using container size as a sink-strength manipulation can be used to empirically test the extent to which growth and *A*~n~ are coordinated.

\
This study utilizes a novel field design to investigate the coordination between growth and *A*~n~ in *Eucalyptus tereticornis* Sm. seedlings, by manipulating container size and thus rooting volume. Seedlings were maintained under well watered conditions in order to isolate the effect of restricted soil volume. We used freely-rooted seedlings as a control for the container size treatments. Empirical results were combined with a simple plant growth model to simulate seedling growth with a *C* mass balance approach, which was then compared to observed harvested seedling mass. The model used whole-plant *C* gain, scaled from instantaneous rates of leaf *A*~n~, to quantify seedling dry mass production over the 120 day experiment.

\
Our hypotheses were as follows: 
\
1). The manipulations of container size were expected to induce a belowground sink limitation compared to free seedlings. We hypothesized that declines in seedling growth would be largest in the smallest containers.

@@

\
*Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were chosen from a single local Cumberland plain cohort. Six additional seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots. Previous container experiments have confirmed that species with tap roots (similar to *E. tereticornis*) use the center of the container as the medium for thick roots leaving the periphery of the soil as the most active sites for fine root proliferation [@biran1980a; @biran1980b]. By using a species with tap root growth and manipulations of container length rather than width, we believed that a more realistic test of growth inhibition through constrained soil volume would be achieved.

\
Six container volumes were used ranging from 5 L to 35 L, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe and were filled with local top soil (described above). Soil in each container was packed to achieve a target soil bulk density that matched local soil conditions of 1.7 g cm^-3^. A Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally planted (free) seedlings. Each experimental block (n=7) contained a complete replicate set of six container volumes as well as one free seedling, with 1 m^2^ spacing. For each free seedling, used as the control, a 1 m^2^ subplot was excavated to the hard layer and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation.

\
Plants were watered weekly or when needed to maintain soil moisture at field capacity (13-15 %). Drain systems were built into each pot to prevent pooling of water throughout the experiment. Pooling of water could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6 mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed. For small containers (5, 10, and 15 L) a simple bottom plug was used to drain excess water from the gravel compartment. Each containers was inspected after every rainfall event to determine if pooling had occurred.

@@

\
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C on a random subset of each container size (n=3) after new leaves were first produced (March 13-14^th^, 2013) and prior to the final harvest (May 14-15~th~, 2013). Each AC~i~ curve was started at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25 &deg;C at saturating light (above). From these curves the photosynthetic parameters, *J*~max~ and *Vc*~max~, were quantified using the biochemical model of [@farquhar1980biochemical] and fit with the 'plantecophys' package [@duursma2015] in R [@RDevelopmentCoreTeam2011] using default parameters.

\
Leaf dark respiration rates (*R*) was measured on each seedling during the same dates as AC~i~ curves. Freshly detached leaves were collected at least 1 hour after sundown and placed inside a conifer chamber attached to the Licor 6400. Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of *R* are standardized rates at 25 &deg;C using a *Q*~10~ value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

@@

\
Leaf total non-structural carbohydrate (TNC) concentration was analyzed on dried and milled leaf samples using a total starch assay kit (Megazyme International, Wicklow, Ireland) and includes the starch  and soluble sugar concentrations (mg g^-1^). Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. TNC-free specific leaf area (SLA~f~, m^2^ kg^-1^) for leaves sampled during gas exchange campaigns, were calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass or the leaf *N* content. Similarly, TNC-free leaf *N* (*N*~f~, %) was calculated on all gas exchange leaves from leaf mass without TNC content.

@@

\
*C*~day~ was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] with the 'plantecophys' package [@duursma2015] in R with the mean photosynthetic parameters (*J*~max~, *Vc*~max~, R and *g*~1~) for each treatment and meteorological data from an onsite weather station. *J*~max~ and *Vc*~max~ were estimated from AC~i~ curves (explained above), *R* was empirically measured and the *g*~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2011reconciling] with observed *g*~s~ values. Methods of the coupled leaf gas exchange model are also described in Duursma et al [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, leaf *A*~n~ rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. *C*~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of *C*~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of *C*~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf *A*~n~ with total plant leaf area. This was accomplished by utilizing 61 previously digitized *Eucalyptus* seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package [@YplantQMC] in R to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Using the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total *A*~n~, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf. The ratio of total *A*~n~ with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedlings, independently for each of the seven soil volume treatments. Next, the linear relationship between $\sigma$~s~ and total leaf area was determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table `r pvetabS`.

\
We then utilized this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf *A*~n~ where sufficient to accurately predict overall seedling biomass production after 120 days. Each model run utilized changes in *A*~n~ and leaf mass fraction (LMF), with values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf *C* gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling *C* was assumed to equal half of the final mass for both modeled and observed seedlings.

\
First, a default model was fitted with mean photosynthetic parameters from the free seedlings and then optimized to produce a LMF which correctly predicted both the leaf mass and total biomass of the harvested free seedling (*M*~free~). This optimized LMF was then used to constrain model runs with soil volume treatment specific *C*~day~, while keeping tissue respiration and turnover parameters constant, to determine if changes in leaf *A*~n~ alone could predict biomass of seedlings in containers (*M*~pots~). Next, model sensitivity to different *C* allocation scenarios, including observed treatment specific LMF and up-regulation of non-leaf tissue respiration by 50 % of default values (*M*~S1~ and *M*~S2~, respectively), was used to test for improvement of model biomass predictions compared to measured harvest biomass. For all cases, seedling biomass production was compared between model output and harvested seedlings with treatment specific mean *C*~day~ by first scaling values to the free seedling control.

@@

\
Total seedling biomass at harvest was significantly different across container volumes (*P* < 0.001) and between container treatments and free seedlings (*P* < 0.001, Table `r pvetab1`). On average, harvested biomass of free seedlings was 84% higher than seedlings in containers. Plant biomass was positively correlated with total leaf area across all treatments (*R*^2^ = 0.97, *P* < 0.001). Differences in biomass partitioning to leaves, stems and roots were not different across soil volumes after variation in seedling biomass across treatments was factored in the analysis (Figure `r allocfig`a,b). Across all treatments, the final harvest root:shoot biomass ratio was conserved in these seedlings which exhibited a slightly higher shoot than root mass ($\overline{x}$ = 0.904, 95% CI = [0.846,1.119]) and the ratio of leaf to fine root mass was also not different (Figure `r allocfig`c).

\
Overall, SRL was higher in seedlings in containers compared to free seedlings but only in some of the container size treatments (Table `r pvetab2`, *P* = 0.009). Fine root length density was significantly higher in the two smallest container sizes and was the lowest in the largest container size (Table `r pvetab2`, *P* < 0.001). Over the duration of the experiment SLA~f~ was higher in free seedlings, but was not different across containers sizes (Table `r pvetab1`, *P* < 0.001) and this pattern was evident beginning in the first gas exchange measurement campaign (*P* < 0.001).

@@

\
Both *J*~max~ and *Vc*~max~ were significantly higher in free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table `r pvetab3`). Leaf dark respiration rates were not significantly different across soil volumes (Table `r pvetab3`). The *g*~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the free seedling treatment and was marginally different across soil volume treatments (Table `r pvetab3`).

\
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although *g*~s~ in free seedlings was generally higher than those in containers (Table `r pvetab3`, *P* < 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not change throughout the course of the experiment. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table `r pvetab1`). Combined these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N at harvest was not different across soil volumes ($\bar{x}$ = 4.5 %), with minimal decreases from pre-planting value ($\bar{x}$ = 4.9 %). This indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

@@

\
Next, we performed a series of model simulations to test possible C allocation scenarios to account for the over predictions of seedling *C* mass. Testing the sensitivity of the model to observed treatment-specific LMF from the final harvest (*M*~S1~), which were each lower than the optimized LMF value (see Figure `r allocfig`a), improved model predictions of seedling C mass but still overestimated seedling total *C* by 32&pm;11.1 % (Figure `r massmodelfig2`a). Using harvest values of LMF, however, does not capture the observed increase in leaf turnover of seedlings in small containers (Figure `r allomfig`c). Thus, the use of harvest LMF values for seedlings in containers in *M*~S1~ likely underestimates daily C allocation to leaves over the final months of the experiment. Increases of 50 % in non-leaf tissue respiration (*M*~S2~) improved biomass estimates slightly but overestimated mass *C* by an average of 46&pm;9.3 % in seedlings with soil volume restriction (Figure `r massmodelfig2`b). With *M*~S2~, non-tissue respiration rates would need to be increased by ca. 250% in order for mass balance to be achieved.

@@

\
Decreases in *A*~sat~ occurred at the same time as reductions in height and diameter of seedlings in containers. This initially suggests a strong link between growth and an apparent down regulation of *A*~n~. However, there are several possible mechanisms that can explain reduced *A*~n~ in small containers including nutrient content, water or reduced sink strength [@poorter2012pot]. It was therefore necessary to examine each of these factors to determine if the induced belowground sink limitation actually triggered photosynthetic down regulation.

>>>>

\
RUNNING HEAD: SINK LIMITATION ALTERS SEEDLING CARBON BALANCE

@@

\
<br/>
<br/>
<br/>
<br/>
In woody species, the coordination of *A*~n~ and growth has been studied with manipulations of *C* source activity. Examples included elevated CO~2~ experiments, for example FACE [reviewed in @ainsworth2005have], and partial defoliation experiments. Elevated CO~2~ has been shown to increase *A*~n~ [@drake1997more; @ainsworth2007response] and across four FACE experiments this resulted in a stimulation of 23 % in forest biomass production [@norby2005forest]. Evidence from a wide range of elevated CO~2~ experiments, however, also reveals that even with an average photosynthetic enhancement of over 30 %, the biomass growth rate only increases by around 10 % [@kirschbaum2011does]. In partial defoliation experiments, increases in *A*~n~ of the remaining foliage are commonly shown, yet are attributed to various mechanisms, including reduction in end product inhibition [@iglesias2002regulation; @zhou2003changes; @handa2005test], enhanced biochemical activity [@ovaska1993b;@layne1995end], increased stomatal conductance [@layne1995end], enhanced leaf nutrient status [@turnbull2007increased] and regulatory sugar signaling [@eyles2013whole]. However, increases in *A*~n~ in defoliation experiments did not always produce increased growth due to reductions in meristem sink strength [@palacio2012fast], *C* limitation to mycorrhizal colonization [@markkola2004defoliation], or an overall decrease in whole plant *C* gain [@ovaska1993a]. These manipulations of C source activity expose unresolved issues with how changes in *A*~n~ do not always infer proportional responses in growth.

\
<br/>
<br/>
<br/>
<br/>
Alternatively, manipulating plant tissue *C* sinks is often used to investigate the correlation between *A*~n~ and growth. This is because metabolic signaling networks, relaying information on *C* and *N* status of different tissues, can regulate photosynthetic activity [@paul2001sink]. If sink inhibition of *A*~n~ occurs, a close coordination between declines in *A*~n~ and growth should be expected. Whether photosynthetic down regulation is evident in woody species has been tested through fruit removal and phloem girdling. In these studies, down regulation of *A*~n~ was frequently correlated to carbohydrate accumulation resulting from reduced tissue sink strength  [@iglesias2002regulation;@urban2007girdling; @haouari2013fruit]. However, reductions in *A*~n~ were also attributed to biochemical limitations prior to carbohydrate accumulation [@nebauer2011photosynthesis], irreversible photo-oxidative damage [@duan2008photosynthetic] and stomatal limitation [@li2005photosynthesis]. These mixed results are not surprising as we still know little about the balance between assimilation, storage and growth across temporal scales in plants [@smith2007coordination]. As these manipulations likely impact source as well as sink activity simultaneously, affect water transport, are very extreme, or are specific to the occurrence of large fruiting sinks, they tell us little about source-sink coordination in typical growing conditions for woody species.

\
<br/>
<br/>
<br/>
<br/>
An alternative experimental approach is to reduce belowground *C* sink strength in tree seedlings by manipulating rooting volume, by varying the container size [@arp1991effects; @nesmith1998effect; @poorter2012pot]. Possible advantages of this approach are that it allows a large range of treatment levels, can be easily compared to naturally planted seedlings and may mimic natural conditions as seedlings compete for space or reach bedrock. Seedlings undergo many physiological and morphological changes in response to rooting volume, including biomass partitioning, *A*~n~, water relations, nutrient uptake and respiration [@nesmith1998effect;@poorter2012pot and references therein]. Inadequate rooting volume may decrease *C* sink strength by progressively restricting root growth in growing plants [@thomas1991root].  Container size studies frequently exhibit photosynthetic down-regulation, likely as a result of sink limitation [@arp1991effects; @mcconnaughay1991physical; @gunderson1994photosynthetic; @sage1994acclimation; @maina2002intra; @ronchi2006growth]. A meta-analysis by Poorter et al. [-@poorter2012pot] concluded that *A*~n~ is the process likely to be the strongest affected by pot size and may best explain the effects on biomass seen in the large number of studies where containers are used. This conclusion arises because plants grown in small containers are shown to accumulate leaf starch while having lower *C* exchange and assimilate export rates [@robbins1988effect]. However, evidence in support for a trade-off between *C* storage and growth in trees is, to date, inconclusive [@palacio2014does]. Based on these previous studies, using container size as a sink-strength manipulation can be used to empirically test the extent to which growth and *A*~n~ are coordinated.

\
<br/>
<br/>
<br/>
<br/>
This study utilizes a novel field design to investigate the coordination between growth and *A*~n~ in *Eucalyptus tereticornis* Sm. seedlings, by manipulating container size and thus rooting volume. Seedlings were maintained under well watered conditions in order to isolate the effect of restricted soil volume. We used freely-rooted seedlings as a control for the container size treatments. Empirical results were combined with a simple plant growth model to simulate seedling growth with a *C* mass balance approach, which was then compared to observed harvested seedling mass. The model used whole-plant *C* gain, scaled from instantaneous rates of leaf *A*~n~, to quantify seedling dry mass production over the 120 day experiment.

\
<br/>
<br/>
<br/>
<br/>
Our hypotheses were as follows: 
\
1). The manipulations of container size were expected to induce a belowground sink limitation compared to free seedlings. We hypothesized that declines in seedling growth would be largest in the smallest containers.

@@

\
<br/>
<br/>
<br/>
<br/>
*Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were chosen from a single local Cumberland plain cohort. Six additional seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots. Previous container experiments have confirmed that species with tap roots (similar to *E. tereticornis*) use the center of the container as the medium for thick roots leaving the periphery of the soil as the most active sites for fine root proliferation [@biran1980a; @biran1980b]. By using a species with tap root growth and manipulations of container length rather than width, we believed that a more realistic test of growth inhibition through constrained soil volume would be achieved.

\
<br/>
<br/>
<br/>
<br/>
Six container volumes were used ranging from 5 L to 35 L, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe and were filled with local top soil (described above). Soil in each container was packed to achieve a target soil bulk density that matched local soil conditions of 1.7 g cm^-3^. A Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally planted (free) seedlings. Each experimental block (n=7) contained a complete replicate set of six container volumes as well as one free seedling, with 1 m^2^ spacing. For each free seedling, used as the control, a 1 m^2^ subplot was excavated to the hard layer and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation.

\
<br/>
<br/>
<br/>
<br/>
Plants were watered weekly or when needed to maintain soil moisture at field capacity (13-15 %). Drain systems were built into each pot to prevent pooling of water throughout the experiment. Pooling of water could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6 mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed. For small containers (5, 10, and 15 L) a simple bottom plug was used to drain excess water from the gravel compartment. Each containers was inspected after every rainfall event to determine if pooling had occurred.

@@

\
<br/>
<br/>
<br/>
<br/>
Photosynthetic CO~2~ response (AC~i~) curves were measured at 25 &deg;C on a random subset of each container size (n=3) after new leaves were first produced (March 13-14^th^, 2013) and prior to the final harvest (May 14-15~th~, 2013). Each AC~i~ curve was started at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25 &deg;C at saturating light (above). From these curves the photosynthetic parameters, *J*~max~ and *Vc*~max~, were quantified using the biochemical model of [@farquhar1980biochemical] and fit with the 'plantecophys' package [@duursma2015] in R [@RDevelopmentCoreTeam2011] using default parameters.

\
<br/>
<br/>
<br/>
<br/>
Leaf dark respiration rates (*R*) was measured on each seedling during the same dates as AC~i~ curves. Freshly detached leaves were collected at least 1 hour after sundown and placed inside a conifer chamber attached to the Licor 6400. Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of *R* are standardized rates at 25 &deg;C using a *Q*~10~ value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

@@

\
<br/>
<br/>
<br/>
<br/>
Leaf total non-structural carbohydrate (TNC) concentration was analyzed on dried and milled leaf samples using a total starch assay kit (Megazyme International, Wicklow, Ireland) and includes the starch  and soluble sugar concentrations (mg g^-1^). Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. TNC-free specific leaf area (SLA~f~, m^2^ kg^-1^) for leaves sampled during gas exchange campaigns, were calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass or the leaf *N* content. Similarly, TNC-free leaf *N* (*N*~f~, %) was calculated on all gas exchange leaves from leaf mass without TNC content.

@@

\
<br/>
<br/>
<br/>
<br/>
*C*~day~ was predicted by using a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] with the 'plantecophys' package [@duursma2015] in R with the mean photosynthetic parameters (*J*~max~, *Vc*~max~, R and *g*~1~) for each treatment and meteorological data from an onsite weather station. *J*~max~ and *Vc*~max~ were estimated from AC~i~ curves (explained above), *R* was empirically measured and the *g*~1~ parameter was generated by fitting the optimal stomatal conductance model from [@medlyn2011reconciling] with observed *g*~s~ values. Methods of the coupled leaf gas exchange model are also described in Duursma et al [-@duursma2014peaked]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15 min intervals, leaf *A*~n~ rates ($\mu$mol CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. *C*~day~ was calculated by converting predicted rates to mass C gain over 15 min time steps (g m^-2^) and then summed for 24 h. This resulted in 120 unique values of *C*~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of *C*~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
<br/>
<br/>
<br/>
<br/>
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf *A*~n~ with total plant leaf area. This was accomplished by utilizing 61 previously digitized *Eucalyptus* seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package [@YplantQMC] in R to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Using the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total *A*~n~, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf. The ratio of total *A*~n~ with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedlings, independently for each of the seven soil volume treatments. Next, the linear relationship between $\sigma$~s~ and total leaf area was determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table `r pvetabS`.

\
<br/>
<br/>
<br/>
<br/>
We then utilized this model to test the hypothesis that the effects of belowground sink limitation on rates of leaf *A*~n~ where sufficient to accurately predict overall seedling biomass production after 120 days. Each model run utilized changes in *A*~n~ and leaf mass fraction (LMF), with values of stem and root respiration rates, to generate total seedling mass and leaf area after 120 days.  Cumulative net leaf *C* gain for each treatment was equal to the sum of each value of C~day,i~ over 120 days and final seedling *C* was assumed to equal half of the final mass for both modeled and observed seedlings.

\
<br/>
<br/>
<br/>
<br/>
First, a default model was fitted with mean photosynthetic parameters from the free seedlings and then optimized to produce a LMF which correctly predicted both the leaf mass and total biomass of the harvested free seedling (*M*~free~). This optimized LMF was then used to constrain model runs with soil volume treatment specific *C*~day~, while keeping tissue respiration and turnover parameters constant, to determine if changes in leaf *A*~n~ alone could predict biomass of seedlings in containers (*M*~pots~). Next, model sensitivity to different *C* allocation scenarios, including observed treatment specific LMF and up-regulation of non-leaf tissue respiration by 50 % of default values (*M*~S1~ and *M*~S2~, respectively), was used to test for improvement of model biomass predictions compared to measured harvest biomass. For all cases, seedling biomass production was compared between model output and harvested seedlings with treatment specific mean *C*~day~ by first scaling values to the free seedling control.

@@

\
<br/>
<br/>
<br/>
<br/>
Total seedling biomass at harvest was significantly different across container volumes (*P* < 0.001) and between container treatments and free seedlings (*P* < 0.001, Table `r pvetab1`). On average, harvested biomass of free seedlings was 84% higher than seedlings in containers. Plant biomass was positively correlated with total leaf area across all treatments (*R*^2^ = 0.97, *P* < 0.001). Differences in biomass partitioning to leaves, stems and roots were not different across soil volumes after variation in seedling biomass across treatments was factored in the analysis (Figure `r allocfig`a,b). Across all treatments, the final harvest root:shoot biomass ratio was conserved in these seedlings which exhibited a slightly higher shoot than root mass ($\overline{x}$ = 0.904, 95% CI = [0.846,1.119]) and the ratio of leaf to fine root mass was also not different (Figure `r allocfig`c).

\
<br/>
<br/>
<br/>
<br/>
Overall, SRL was higher in seedlings in containers compared to free seedlings but only in some of the container size treatments (Table `r pvetab2`, *P* = 0.009). Fine root length density was significantly higher in the two smallest container sizes and was the lowest in the largest container size (Table `r pvetab2`, *P* < 0.001). Over the duration of the experiment SLA~f~ was higher in free seedlings, but was not different across containers sizes (Table `r pvetab1`, *P* < 0.001) and this pattern was evident beginning in the first gas exchange measurement campaign (*P* < 0.001).

@@

\
<br/>
<br/>
<br/>
<br/>
Both *J*~max~ and *Vc*~max~ were significantly higher in free seedlings (30 % and 26 %, respectively) than container-grown seedlings with little variation between container volume treatments (Table `r pvetab3`). Leaf dark respiration rates were not significantly different across soil volumes (Table `r pvetab3`). The *g*~1~ parameter, generated for each seedling from the Medlyn et al [-@medlyn2011reconciling] optimal stomatal conductance model, was lowest in the free seedling treatment and was marginally different across soil volume treatments (Table `r pvetab3`).

\
<br/>
<br/>
<br/>
<br/>
Neither $\Psi$~pd~ nor $\Psi$~l~ were different across treatments, with mean values of -0.27 and -1.2 MPa across all seedlings, respectively. Although *g*~s~ in free seedlings was generally higher than those in containers (Table `r pvetab3`, *P* < 0.001), the mean rates for all seedlings were high at 0.37 mol m^-2^ s^-1^ and did not change throughout the course of the experiment. Additionally, leaf $\delta$^13^C at final harvest was not different across treatments (Table `r pvetab1`). Combined these indices provide strong evidence that water stress was not apparent on these seedlings throughout the experiment. Soil N at harvest was not different across soil volumes ($\bar{x}$ = 4.5 %), with minimal decreases from pre-planting value ($\bar{x}$ = 4.9 %). This indicates that nutrient leaching from free seedlings or from draining of containers following natural rainfall events did not differ between treatments.

@@

\
<br/>
<br/>
<br/>
<br/>
Next, we performed a series of model simulations to test possible C allocation scenarios to account for the over predictions of seedling *C* mass. Testing the sensitivity of the model to observed treatment-specific LMF from the final harvest (*M*~S1~), which were each lower than the optimized LMF value (see Figure `r allocfig`a), improved model predictions of seedling C mass but still overestimated seedling total *C* by 32&pm;11.1 % (Figure `r massmodelfig2`a). Using harvest values of LMF, however, does not capture the observed increase in leaf turnover of seedlings in small containers (Figure `r allomfig`c). Thus, the use of harvest LMF values for seedlings in containers in *M*~S1~ likely underestimates daily C allocation to leaves over the final months of the experiment. Increases of 50 % in non-leaf tissue respiration (*M*~S2~) improved biomass estimates slightly but overestimated mass *C* by an average of 46&pm;9.3 % in seedlings with soil volume restriction (Figure `r massmodelfig2`b). With *M*~S2~, non-tissue respiration rates would need to be increased by ca. 250% in order for mass balance to be achieved.

@@

\
<br/>
<br/>
<br/>
<br/>
Decreases in *A*~sat~ occurred at the same time as reductions in height and diameter of seedlings in containers. This initially suggests a strong link between growth and an apparent down regulation of *A*~n~. However, there are several possible mechanisms that can explain reduced *A*~n~ in small containers including nutrient content, water or reduced sink strength [@poorter2012pot]. It was therefore necessary to examine each of these factors to determine if the induced belowground sink limitation actually triggered photosynthetic down regulation.

\
<br/>
<br/>
<br/>
<br/>
With high rates of *g*~s~, non-limiting leaf water potential ($\Psi$~pd~) and consistent leaf $\delta$^13^C across soil volume treatments there was little evidence that water stress caused the reduction in *A*~n~. This finding is consistent with other container size studies without drought treatments.  For example, reduced *A*~max~ in cotton seedlings grown at elevated CO~2~ was attributed to sink-limited feedback inhibition from inadequate rooting volume, not decreased *g*~s~ [@thomas1991root]. Additionally, severe reductions in *A*~n~ in coffee plants were not attributed to impacts of container size on leaf water potentials or *g*~s~ [@ronchi2006growth]. It is likely that reductions in *A*~n~ of well-watered seedlings observed in our study of *E.tereticornis* seedlings was instead the result of limiting soil nutrients or space restriction on belowground sink strength.
